---
title: "County-Level Chronic Disease and Air Quality Analysis"
subtitle: "CDC PLACES + EPA PM2.5 — Reproducible Pipeline Demo"
author: "Martin Donovan"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: flatly
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(DBI)
library(RPostgres)
library(tidyverse)
library(gt)
library(config)

cfg <- config::get()

con <- dbConnect(
  RPostgres::Postgres(),
  host     = cfg$db$host,
  port     = cfg$db$port,
  dbname   = cfg$db$dbname,
  user     = cfg$db$user,
  password = Sys.getenv("DB_PASSWORD_EPI")
)

# Load mart — all analysis draws from here
mart <- dbGetQuery(con, "SELECT * FROM marts.mart_epi_analysis")
dbDisconnect(con)
```

## Overview

This report summarizes county-level chronic disease prevalence (CDC PLACES) and
ambient PM2.5 air quality (EPA AQS) across the United States. Data flows from raw
public sources through a dbt transformation pipeline to produce the analysis table
underlying this report.

**Data sources:**

- CDC PLACES County Data — chronic disease and behavioral risk factor prevalence
- EPA AirData Annual PM2.5 Summary — ambient fine particulate matter by monitoring station

**Pipeline:** Raw CSV → R ingestion → PostgreSQL raw schema → dbt staging/intermediate/mart → this report

---

## Data Summary

```{r summary-table}
# TODO: build a gt summary table showing:
# - Number of counties covered
# - Year range
# - % counties with PM2.5 data
# - Key measures available

# Placeholder
mart %>%
  summarise(
    counties = n_distinct(county_fips),
    states = n_distinct(state_abbr),
    years = paste(min(year, na.rm=TRUE), max(year, na.rm=TRUE), sep="-"),
    pct_with_pm25 = round(mean(!missing_pm25, na.rm=TRUE) * 100, 1)
  ) %>%
  gt() %>%
  cols_label(
    counties = "Counties",
    states = "States",
    years = "Years",
    pct_with_pm25 = "% with PM2.5 data"
  )
```

---

## Chronic Disease Prevalence

```{r prevalence-table}
# TODO: gt table showing national median + IQR for each PLACES measure
# Rows: measure | median | IQR | n counties
# Use confidence intervals from mart where available
```

---

## PM2.5 Distribution

```{r pm25-distribution}
# TODO: histogram or density plot of county-level PM2.5 annual mean
# Color by census region or urban/rural if USDA data added later
# ggplot(mart %>% filter(!is.na(pm25_annual_mean))) +
#   geom_histogram(aes(x = pm25_annual_mean), bins = 40) + ...
```

---

## Exposure-Outcome Relationship

PM2.5 annual mean vs. chronic disease prevalence across counties.

```{r scatter-pm25-outcome}
# TODO: scatter plot — x = pm25_annual_mean, y = obesity_pct (or chosen outcome)
# Color by state or region
# Add simple lm() trend line
# Note: this is descriptive, not causal inference
```

---

## County-Level Summary Table

```{r county-table}
# TODO: searchable/sortable gt or DT table of mart_epi_analysis
# Columns: state, county, year, pm25, key measures
# Filter to most recent year for default view
```

---

## Methods Note

This analysis uses county-level aggregate data. No individual-level data is used.
All data sources are publicly available. EPA PM2.5 values are annual means aggregated
across monitoring stations within each county; counties with fewer than 75% valid
observations are excluded. CDC PLACES values are age-adjusted prevalence estimates.
This pipeline is a demonstration of reproducible data infrastructure; statistical
conclusions require appropriate epidemiologic methods not applied here.
