---
title: "County-Level Chronic Disease and Air Quality Analysis"
subtitle: "CDC PLACES + EPA PM2.5 — Reproducible Pipeline Demo"
author: "Martin Donovan"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: flatly
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(DBI)
library(RPostgres)
library(tidyverse)
library(gt)
library(config)

cfg <- config::get()

con <- dbConnect(
  RPostgres::Postgres(),
  host     = cfg$db$host,
  port     = cfg$db$port,
  dbname   = cfg$db$dbname,
  user     = cfg$db$user,
  password = Sys.getenv("DB_PASSWORD_EPI")
)

# Load mart — all analysis draws from here
mart <- dbGetQuery(con, "SELECT * FROM marts.mart_epi_analysis")
dbDisconnect(con)
```

## Overview

This report summarizes county-level chronic disease prevalence (CDC PLACES) and
ambient PM2.5 air quality (EPA AQS) across the United States. Data flows from raw
public sources through a dbt transformation pipeline to produce the analysis table
underlying this report.

**Data sources:**

- CDC PLACES County Data — chronic disease and behavioral risk factor prevalence
- EPA AirData Annual PM2.5 Summary — ambient fine particulate matter by monitoring station

**Pipeline:** Raw CSV → R ingestion → PostgreSQL raw schema → dbt staging/intermediate/mart → this report

---

## Data Summary

```{r summary-table}
# [PART 11 · STEP 11.1] Data summary gt table
# Build a gt table showing:
#   - Number of counties covered      n_distinct(county_fips)
#   - Number of states                n_distinct(state_abbr)
#   - Year range                      min(year) – max(year)
#   - % counties with PM2.5 data      mean(!missing_pm25) * 100
#
# Use summarise() then gt() with cols_label() for readable column names
```

---

## Chronic Disease Prevalence

```{r prevalence-table}
# [PART 11 · STEP 11.2] National prevalence summary table
# Reshape mart from wide to long with pivot_longer() across the *_pct columns
# Summarise by measure: median, Q1, Q3, n counties with data
# Display with gt() — one row per measure
```

---

## PM2.5 Distribution

```{r pm25-distribution}
# [PART 11 · STEP 11.3] Histogram of county-level PM2.5
# Filter to !is.na(pm25_annual_mean)
# ggplot geom_histogram() or geom_density()
# Optional: color/facet by census region if you add a region column later
```

---

## Exposure-Outcome Relationship

PM2.5 annual mean vs. chronic disease prevalence across counties.

```{r scatter-pm25-outcome}
# [PART 11 · STEP 11.4] Scatter plot — PM2.5 vs. chosen outcome
# x = pm25_annual_mean, y = obesity_pct (or another measure)
# geom_point(alpha = 0.3) + geom_smooth(method = "lm")
# Color points by state or leave grey
# Note in caption: descriptive association only, not causal
```

---

## County-Level Summary Table

```{r county-table}
# [PART 11 · STEP 11.5] Searchable county data table
# Filter mart to most recent year for default view
# Columns: state_abbr, county_name, year, pm25_annual_mean,
#          obesity_pct, diabetes_pct, smoking_pct
# Use DT::datatable() for searchable/sortable output
# Or gt() for a formatted static table
```

---

## Methods Note

This analysis uses county-level aggregate data. No individual-level data is used.
All data sources are publicly available. EPA PM2.5 values are annual means aggregated
across monitoring stations within each county; counties with fewer than 75% valid
observations are excluded. CDC PLACES values are age-adjusted prevalence estimates.
This pipeline is a demonstration of reproducible data infrastructure; statistical
conclusions require appropriate epidemiologic methods not applied here.
